---
title: "Block III"
author: "Sara Dovalo and Javier Muñoz Flores"
date: "21/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(h2o)
library(dplyr)
library(ggplot2)
```

# Introduction

The content of this section is, mainly, a further description of the dataset used as well as the exposition of the problem to solve. 

The dataset selected contains several patient records of different medical measurements in order to predict wether a person is more likely to suffer a heart disease, i.e. a failure . The data has been retrieved from the public [*kaggle*](https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction) repository and it is the product of the combination of five different datasets from different regions of EEUU. The final dataset contains in total 918 instances and 12 attributes, which 7 of them are categorical and the five remaining are numerical:

 - `Age`(*quantitative*): age of the patient in years
 - `Sex`(*qualitative*): sex of the patient [*M*: Male, *F*: Female]
 - `ChestPainType`(*qualitative*): Angina type, i.e. chest pain, frequently caused when the heart muscle is not able to get enough oxygen-rich blood [*TA*: Typical Angina, *ATA*: Atypical Angina, NAP: Non-Anginal Pain, ASY: Asymptomatic]
 - `RestingBP`(*quantitative*): resting blood pressure [mm Hg]. A normal level is less than 180 mm Hg.
 - `Cholesterol`(*quantitative*): serum cholesterol [mm/dl]
 - `FastingBS`(*qualitative*): fasting blood sugar [1: if FastingBS > 120 mg/dl, 0: otherwise]
 - `RestingECG`(*qualitative*): resting electrocardiogram results [Normal: Normal, ST: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV), LVH: showing probable or definite left ventricular hypertrophy by Estes' criteria]
 - `MaxHR`(*quantitative*): maximum heart rate achieved [Numeric value between 60 and 202]
 - `ExerciseAngina`: exercise-induced angina [Y: Yes, N: No]
 - `Oldpeak`(*quantitative*): oldpeak = ST [Numeric value measured in depression]
 - `ST_Slope`(*qualitative*): the slope of the peak exercise ST segment [Up: upsloping, Flat: flat, Down: downsloping]
 - `HeartDisease`(*qualitative*): class variable [1: heart disease, 0: Normal]

Clearly, it is a binary classification problem since the target variable has two levels.

# Preprocessing 

First of all , it is suitable to visualize the data and to identify if there are missing values which could add noise and disrupt the performance of the future models created.

```{r, include = FALSE}
# Load data and rename first column
heart.data <- read.csv("heart.csv")
# Data summary
summary(heart.data)
```

```{r}
# Count number of zeros (missing values)
heart.data %>% filter(Cholesterol == 0) %>% summarize(count = n())
```

We notice that the variable `Cholesterol` contains 172 values equal to zero. They must be considered as missing values as a person cannot that such a low level of cholesterol in blood. We decide to eliminate those observations that have a zero in that variable. 

```{r}
# Eliminate rows which contain 0 in variable Cholesterol
heart.data = heart.data %>%
filter(Cholesterol != 0)
# Number of missing values
heart.data %>% filter(Cholesterol == 0) %>% summarize(count = n())
```


Furthermore, since the dataset includes several categorical attributes, it is necessary to transform them into *factors*. 

```{r}
# Categorical variables as factors
heart.data = heart.data %>% 
       mutate_each_(funs(factor(.)),c(2,3,6,7,9,11,12))
str(heart.data)
```

To finish this section, we check if the classes of the response `HeartDisease` are balanced or not. We visualize it through a barplot to see it more clearly.

```{r}
# Count the observation of each class
heart.data %>%
count(HeartDisease)
# Barplot
ggplot(heart.data, aes(HeartDisease)) + geom_bar()
```
The classes are balanced, it will be not needed to over/undersampling the sample.

## Modelling with `H2O`package

We follow the same procedure that we did in classroom for using `h2o.automl()`, i.e. fitting, benchmarking, predicting and explaining, in that order.

### Fitting 

Before starting with the selection of the models, the *local H2O cluster* is initialized in order to leverage the parallelization in the virtual machine which the package provides and to use H2O functions.

The first step is to convert the data into a `h2o`object and then, to identify the names of the predictors as well as the name of the response.

Then, it is important to mention that there will not be splitting into train and test, since we will use cross-validation metrics on the leaderboard model. Thus, we have to specify the number of folds (in our case will be `nfolds = 8`) needed to the cross-validation process. However, we do not have to indicate the predictors names, instead only the`dataFrame`and the response variable. In addition, we do not exclude any algorithm in `h2o.automl()` method, but we limit the time for fitting the models (`max_runtime_secs = 30` and `max_runtime_secs_per_model = 5`)

```{r}
# Initialize h2o
h2o.init()
# Send data to local H2O cluster
data.h <- as.h2o(heart.data)
# Data summary 
h2o.describe(data.h)
# Identify the names of the response and predictors
resp_h <- "HeartDisease"
pred_h<- setdiff(names(data.h), resp_h)
# Call h2o.automl()
model_h <- h2o.automl(y = resp_h, training_frame = data.h, max_runtime_secs = 30,max_runtime_secs_per_model = 5, nfolds = 8,
seed = 1, verbosity = NULL)
```

### Benchmarking

In this section, we have to explore the leaderboard of models in order to carry out a first comparison.

```{r}
# Leaderboard
lead_h <- model_h@leaderboard
names(lead_h)[5] <- "mpce" # Rename mean_per_class_error to shorten output
print(lead_h[, -6], n = nrow(lead_h)) # Exclude final column to fit the table in one page
```
The leader has been a *StackedEnsemble* model with an error

### Prediction

We select the leader of the models and predict a few rows.

```{r}
h2o.predict(object = model_h@leader, newdata = data.h[1:8, ])
```

### Explanation

```{r}
exp <- h2o.explain(object = model_h, newdata = data.h)
exp
```






